---
title: "bayesian_project"
output: html_document
author: Phoebe Teskey
email: phoebeteskey@arizona.edu
date: 15/04/2026
---

```{r}
packages.needed<-c("ggplot2","bayesplot","fitdistrplus","reshape2",

"qqplotr","gridExtra","ggeffects", "GGally","broom","qqplotr","brms",

"bayesplot","doBy")

lapply(packages.needed, FUN = require, character.only = T)
```


```{r}
veg <- read.csv("../data/soil_transects.csv")

head(veg)

```

Question: How does soil type affect plant abundance?

```{r}
# convert to factors
veg$SpeciesName <- as.factor(veg$SpeciesName)
veg$SiteID <- as.factor(veg$SiteID)
veg$Transect <- as.factor(veg$Transect)
veg$SoilType <- as.factor(veg$SoilType)

str(veg$nValue)
veg$nValue <- as.integer(as.numeric(veg$nValue))
veg <- veg %>%
  na.omit(nValue, SoilType, ESD_Designation, SiteID, Transect, SpeciesName)

```

```{r}
get_prior(formula=nValue~SoilType, data = veg, family = negbinomial())
```
When a prior is "flat" that means it has no prior information at all. 

```{r}
prior.m2 = c(set_prior("normal(0,2)", class = "Intercept"),
set_prior("normal(0,1)", class = "b"),
set_prior("exponential(1)", class = "sd"),
set_prior("exponential(1)", class = "shape"))
```

Weakly informative priors are set. Now let's run the model.
Setting priors is a bit of an art. It requires knowledge of the system.

```{r}
m2.brms=brm(formula = nValue ~ SoilType + ESD_Designation + (1 | SiteID/Transect), data = veg, family = negbinomial(),
            prior = prior.m2, chains = 4, cores = 4,
            iter = 2000, seed = 123)
summary(m2.brms)
```

Rhat = 1, which means chains converged, which is good.
Intercept = 2.41, log transformed is 11.1 plants per sample quadrat.
Soil type credible intervals include 0, which tells you that soil type doesn't really affect plant abundance.
When looking at the plant types, we can see that there's a strong negative effect when annual forbs are present. So there's less plant abundance where annual forbs are present.
When introduced grass is present, those areas have a higher plant abundance. 
shape = 1, indicating that variance is greater than the mean, which means negative binomial is the correct family for this dataset.

Results:
Vegetation type was a stronger predictor of plant abundance than soil type. Introduced grass presence is associated with higher plant abundance, while annual forbs are associated with lower plant abundance.

```{r}
install.packages("usethis")
library(usethis)
use_git()
use_github()
```

